Staged Abstract Interpreters

### 2019-03-23 paper/10_intro.tex
revise

\paragraph{Abstraction, Abstraction, and Abstraction}
To enable such flexibility between different semantics, we combine generic 
programming and generative programming, and extensively use abstract type members 
over multiple abstractions, which can be later instantiated.

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form
  \cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95,
  DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}, by plugging
  different monads, the underlying semantics of the interpreter is changed.
  \todo{mention that direct-style also works?}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values declared in the generic interpreter are
  abstract types; we also define several abstract methods that manipulate
  environments or stores.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\end{itemize}

@@

@@

\paragraph{The paper is organized as follows:}

\begin{itemize}
  \item We progressively present a unifying construction that naturally extends
    the first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a small functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation (Section~\ref{unstaged_conc}). Then we lift it to the staged
    version by replacing the binding-time types (Section~\ref{stagedinterp}); and to
    the abstract interpreter by applying abstraction to the environment, store, and
    values (Section~\ref{unstaged_abs}). Finally, we show the combination of the two
    transformations, dubbed \textit{staged abstract interpreters}, can be easily
    derived (Section~\ref{sai}).

@@

  \item We conduct two case studies to show the applicability and usefulness
    of applying meta-programming to static analysis (Section~\ref{cases_study}).
    (1) We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} technique (Section~\ref{cs_ac}).
    Our approach and AC are able to achieve the same goal, however, as we show,
    one of the advantage of staging through types is it does not require
    whole-analyzer refactoring.
    (2) We extend the staged abstract interpreter to two variants of
    control-flow analysis: context-sensitive pushdown control-flow analysis and
    \todo{precise-store?} (Section~\ref{sai}).

>>>>

@@

To enable such flexibility between different semantics, we combine generic 
programming and generative programming, and extensively use abstract type members 
over multiple abstractions, which can be later instantiated.

\begin{itemize}
\item Monadic abstractions, which encapsulate the following two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed.
  \todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\item Env-store-value abstractions, which control whether the interpreter
  behaves as a concrete interpreter or an abstract interpreter.
  The environments, stores and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\end{itemize}

@@

\paragraph{Contributions} The contribution of the paper is twofold:
1) Intellectually, we present a unifying construction that naturally extends the first 
   Futamura projection to abstract interpreters. 
2) Pragmatically, staging abstract interpreters is an approach to improve performance 
   while untouch the soundness. We study the applicability and evaluate the performance 
   of this approach on various analyses.

The paper is organized as follows:

\begin{itemize}
  \item Starting from a generic interface of the interpreter for a functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation (Section~\ref{unstaged_conc}). Then we lift it to the staged
    version by replacing the binding-time types (Section~\ref{stagedinterp}); and to
    the abstract interpreter by applying abstraction to the environment, store, and
    values (Section~\ref{unstaged_abs}). Finally, we show the combination of the two
    transformations, dubbed \textit{staged abstract interpreters}, can be readily
    derived (Section~\ref{sai}).

@@

  \item We conduct two case studies to show the applicability and usefulness
    of applying meta-programming to static analysis (Section~\ref{cases_study}).
    (1) We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} technique (Section~\ref{cs_ac}).
    Our approach and AC are able to achieve the same goal, however, as we show,
    one of the advantage of staging through types is it does not require
    whole-analyzer refactoring.
    (2) We extend the staged abstract interpreter to two variants of
    control-flow analysis: context-sensitive pushdown control-flow analysis and
    \todo{store-widened?} (Section~\ref{sai}).

### 2019-03-23 paper/20_bg.tex
revise

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interface shared between the four different
interpreters. We first instantiate the interpreter to the concrete one; then we
proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}), by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs});

\input{21_bg_interp}
\input{23_bg_stage}
\input{22_bg_abs}

>>>>

In this section, we first describe the abstract syntax of the language we will
analyze, then present the generic interpreter shared among the four different
semantics, after which, we instantiate the interpreter to the concrete one. Then we
will proceed in two directions: by staging it we obtain a staged concrete interpreter
(Section~\ref{stagedinterp}); by abstracting it we obtain an unstaged abstract
interpreter (Section~\ref{unstaged_abs}).

We choose to use Scala and monad to demonstrate the idea, but the approach
is not restricted to our choice. One can use imperative style and other MSP languages 
(such as MetaOCaml \cite{DBLP:conf/gpce/CalcagnoTHL03, DBLP:conf/flops/Kiselyov14} 
or Template Haskell \cite{Sheard:2002:TMH:636517.636528} \todo{cite the recent one})
to construct such staged abstract interpreters.

\input{21_bg_interp}
\input{22_bg_stage}
\input{23_bg_abs}

### 2019-03-23 paper/21_bg_interp.tex
revise

Monads provide an elegant and extensible way to build
interpreters\cite{DBLP:conf/popl/LiangHJ95, Steele:1994:BIC:174675.178068,
Wadler:1992:EFP:143165.143169}, and recently the approach has been extended to
abstract interpreters \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17}. In section, with the multi-stage programming
and abstract interpretation in mind, we use moandic operations to construct an
generic interface in the style of big-step definitional interpreter.

>>>>

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
takes a monad as argument and produces another monad. By using monad transformers, 
we can combine multiple monads into a single one.
Constructing extensible interpreters using monad transformers was first proposed by \citet{DBLP:conf/popl/LiangHJ95},
and later has been applied to abstract interpreters \cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.
In section, with the multi-stage programming and monad transformers in mind
while leave them as abstract type members, we present the generic interface of
a big-step definitional interpreter.

\todo{}
